---
title: "Reproducible Research Project 1"
author: "Rich Huebner"
date: "January 22, 2016"
output: html_document
---

# Introduction
This project uses data from a personal activity monitoring device. The data contains information about the number of steps a person took at 5-minute intervals throughout the day. Data was collected anonymously during October & November, 2012.

This document outlines all of the procedures used to determine the results of the questions in the assignment.

# The data set

This data set was given as part of the Data Science course: Reproducible Research.

The data set can be downloaded here: 
[https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip](Data Set)


# Setting up the environment

```{r, echo=TRUE}
library(knitr)
opts_chunk$set(echo=TRUE)
```

First, load the libraries.
```{r}
library(ggplot2)
library(dplyr)
library(knitr)
library(lubridate)
```

Change to my local working directory. I have everything on a D: under various directories. Change this to your own.
```{r}
setwd("D:\\data\\Rprojects\\RResearch_Project1")
```

Read data set first and tidy up the date data
```{r}
data <- read.csv("activity.csv", header=TRUE, sep=",")
data$date <- ymd(data$date)
```

## Question 1: What is the mean total number of steps taken per day?
It is OK to ignore missing values in the data set.

1. Calculate the total number of steps taken per day.

```{r}
steps <- data %>%
     filter(!is.na(steps)) %>%
     group_by(date) %>%
     summarize(steps = sum(steps))

```

2. Make a histogram of the total number of steps taken per day.

```{r}
ggplot(steps, aes(x=steps)) + 
     geom_histogram(fill="navyblue") + labs(x="# of Steps", y="Frequency")

```

3. Calculate and report on the mean and median of the total number of steps taken per day.

```{r}
avg_steps <- mean(steps$steps, na.rm=TRUE)
median_steps <- median(steps$steps, na.rm=TRUE)
max_steps <- max(steps$steps, na.rm=TRUE)

avg_steps
median_steps
```


## Question 2: What is the average daily activity pattern?

Create a time series plot (i.e. type = "l) of the 5-minute interval (x-axis) and the average number
of steps taken, averaged across all days. 

1. First create the interval.

```{r}

interval <- data %>%
          filter(!is.na(steps)) %>%
          group_by(interval) %>%
          summarize(steps = mean(steps))
```

2. Create the plot now.

```{r}

ggplot(interval, aes(x=interval, y=steps)) +
     geom_line(color="navyblue")

```

3. Which 5-minue interval that, on average, contains the maximum number of steps

```{r}
i <- interval[which.max(interval$steps), ]
i
```




## Question 3: Imputing missing values.
Calculate and report the total number of missing values (coded as NA). The presense of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the data set. ()

```{r}
sum(is.na(data$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset. 
3. Create a new dataset that is equal to the original data set but with the missing data filled in.

I use the average number of steps within the same 5-minute interval to impute the missing values.
I also combined a few steps into this next section of code.

```{r}
imputed_data <- data
na_values <- is.na(imputed_data$steps)
avg_i <- tapply(imputed_data$steps, imputed_data$interval, mean, na.rm=TRUE, simplify = TRUE)

imputed_data$steps[na_values] <- avg_i[as.character(imputed_data$interval[na_values])]

imputed_data$steps[na_values] <- tapply(imputed_data$steps, imputed_data$interval, mean, na.rm=TRUE, simplify=TRUE)

```

Check to make sure the NAs are all gone.
```{r}

sum(is.na(imputed_data$steps))
```


4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

First get the total # of steps taken each day from the imputed data set.

```{r}
all_steps <- imputed_data %>%
               filter(!is.na(steps)) %>%
               group_by(date) %>%
               summarize(steps = sum(steps)) %>%
               print
```

Then make a histogram of the total number of steps taken each day after missing values are imputed.
I tend to go with the ggplot2 package for my plots wherever possible.

```{r}

ggplot(all_steps, aes(x=steps)) +
          geom_histogram(fill="navyblue") +
          labs(title="Histogram of Steps/Day, Imputed Data Set", x="Steps/day", y="Frequency")

```

Display the mean and then the median from the imputed dataset.
```{r}
imean <- mean(all_steps$steps, na.rm=TRUE)
imed <- median(all_steps$steps, na.rm=TRUE)

imean
imed
```

Note: The impact of imputing the data results in a mean and median of the same value: 10766.19 steps.



# Question 4: Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
2. Make a panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

Here's the new factor variable:

```{r}
final <- mutate(imputed_data, weektype= ifelse(weekdays(imputed_data$date) == "Saturday" | weekdays(imputed_data$date) == "Sunday", "weekend", "weekday"))
final$weektype <- as.factor(final$weektype)
head(final)

```

Determine the mean # of steps taken per 5-minute interval across weekdays and weekends.

```{r}
avg_final <- final %>%
          group_by(interval, weektype) %>%
          summarize(steps = mean(steps))

```

...and make the plot using ggplot!

```{r}
ggplot(avg_final, aes(interval, steps, color = weektype)) +
          geom_line(color="red1") +
          facet_wrap(~weektype, ncol=1, nrow=2) +
          labs(title="Weekday/Weekend Differences during various Intervals", x="Interval", y="# of Steps")

```

From the results of this graph, it appears as though there are some distinct differences between weekday and
weekend activity. During the week, this person looks like they go out for some activity in the morning, and then 
the remainder of the day has much less activity.  Most of the activity occurrs on the weekend days.


.. and here's some interesting information not required within the assignment...

The mean # of steps taken for a weekday and the a weekend, for a 5-minute interval..

```{r}
aggregate(steps ~ weektype, final, mean)

```


This concludes the data analysis.









